Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-04-14 analysis/paper/paper.Rmd
CHANGE COLOR PALETTE TO YlGnBu
SKH suggested finding a better color palette for the example data record in figure. YlGnBu from colorbrewer provides better visualization for the lower end of value range.

p <- ggplot(d,aes(day,hour,fill=percent_dry))+
  geom_tile(color= "white",size=0) + 
  # scale_fill_gradientn(colors = viridisLite::turbo(n=256),
  #                      name = "hourly percent dry", aesthetics = "fill",
  #   guide = guide_colorbar(title.position = "bottom", barwidth = 15, 
  #                          barheight = 0.5, title.hjust = 0.5)
  #   )
  scale_fill_viridis_c(name = "hourly percent dry", aesthetics = "fill",
    guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                           barheight = 0.5, title.hjust = 0.5)
  )
p <- p + facet_grid(year~month)
#p <- p + scale_y_continuous(trans = "reverse")
p <- p + scale_x_continuous(breaks = c(1,10,20,30))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  ggtitle("Observation records are incomplete because not all data is transmitted",
          subtitle = "percent-dry data for a single ribbon seal across two years") +
  xlab("day of month") + ylab("hour (UTC)")
p
```

@@

In addition to sex and age class (non-dependent *young-of-the-year*, sexually immature
*subadults*, and mature *adults*), we gathered additional variables that might help
explain variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and local solar hour (for diurnal effects). Solar hour was calculated using the
{solaR} package [@perpinan2012jss] within the R statistical environment [@rcoreteam2019]
using the daily locations. We also compiled several meteorological variables shown to
affect haul-out behavior in other Arctic pinnipeds [@reder2003pb; @udevitz2009cjz;
@perry2017]. In particular, we linked locations from satellite tags to key weather values
from the North American Regional Reanalysis (NARR) model produced by the National Centers
for Environmental Prediction [@mesinger2006bams]. The NARR model assimilates observational
data to produce a long-term picture of weather over North America. Numerous weather
variables are made available across the region 8 times daily. For this study, NARR weather
values were subset to the extent of our study area over the Bering and Chukchi Seas at
3-hr intervals based on the native grid resolution of 32 km (1024 sq. km). The following
meteorological variables were interpolated and assigned to seal locations using a bilinear
method: 1) air temperature at 2m above the earth's surface, 2) wind consisting of
northerly and easterly vector components converted to wind speed using the Euclidean norm,
3) barometric pressure at sea level, and 4) precipitation (Table
\@ref(tab:tblCovariates)).

@@

To properly account for temporal dependence within a computationally tractable framework,
we used generalized linear mixed pseudo-models (GLMPMs; [@verhoef2010cs]) to model
variation in haul-out behavior as a function of (1) covariate predictors, (2) temporally
autocorrelated random effects, and (3) individual random effects representing
heterogeneity in individual behavior. We used the {glmmLDTS} package [@verhoef2010cs] to
implement GLMPMs in the R programming environment [@rcoreteam2021]. Owing to the large
number of records, we fit separate models to bearded, ribbon, and spotted seal data sets.
We fit two different models to our datasets. First, for each species, we fitted a
year-independent model that predicted average haul-out behavior as a function of
demographic, environmental, seasonal, and diurnal effects. Second, for ribbon and spotted
seals (which had considerably more data than bearded seals), we fitted models that
included all the effects from the first model, but also permitted annual variation in
haul-out timing. This second class of models was used to examine whether haul-out patterns
varied by year and to determine the annual timing of apparent peaks in haul-out behavior.

@@

The second set of models that included annual variation in haul-out patterns uncovered
significant contributions for linear and quadratic interactions between day and year in
both ribbon seals (day $\times$ year,
$F_{`r paste(ribbon_year_type3$day_year$Num.df, ribbon_year_type3$day_year$Den.df, sep=",")`}$
= `r ribbon_year_type3$day_year$F.value`; $p$ = `r ribbon_year_type3$day_year$Prob.F`;
day^2^:year,
$F_{`r paste(ribbon_year_type3$day2_year$Num.df, ribbon_year_type3$day2_year$Den.df, sep=",")`}$
= `r ribbon_year_type3$day2_year$F.value`; $p$ = `r ribbon_year_type3$day2_year$Prob.F`)
and spotted seals (day $\times$ year,
$F_{`r paste(spotted_year_type3$day_year$Num.df, spotted_year_type3$day_year$Den.df, sep=",")`}$
= `r spotted_year_type3$day_year$F.value`; $p$ = `r spotted_year_type3$day_year$Prob.F`;
day^2^:year,
$F_{`r paste(spotted_year_type3$day2_year$Num.df, spotted_year_type3$day2_year$Den.df, sep=",")`}$
= `r spotted_year_type3$day2_year$F.value`; $p$ =
`r spotted_year_type3$day2_year$Prob.F`). Predicted distributions of haul-out activity
were largely unimodal, but varied some among and within years with respect to the timing
and magnitude of haul-out peaks (Figure \@ref(fig:annualHO)). It is important to note that
predicted variation in annual haul-out patterns likely reflects both process error and
sampling variability. While we did remove any years where only one deployment in a species
+ age-sex group was present, there are still some years where the pattern shown is
informed by a small number of tags that may not represent population level processes.

>>>>

p <- ggplot(d,aes(day,hour,fill=percent_dry))+
  geom_tile(color= "white",size=0) + 
  scale_fill_distiller(palette = "YlGnBu",
                       name = "hourly percent dry", aesthetics = "fill",
    guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                           barheight = 0.5, title.hjust = 0.5)
    )
  # scale_fill_viridis_c(name = "hourly percent dry", aesthetics = "fill",
  #   guide = guide_colorbar(title.position = "bottom", barwidth = 15,
  #                          barheight = 0.5, title.hjust = 0.5)
  # )
p <- p + facet_grid(year~month)
#p <- p + scale_y_continuous(trans = "reverse")
p <- p + scale_x_continuous(breaks = c(1,10,20,30))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  ggtitle("Observation records are incomplete because not all data is transmitted",
          subtitle = "percent-dry data for a single ribbon seal across two years") +
  xlab("day of month") + ylab("hour (UTC)")
p
```

@@

In addition to sex and age class, we gathered additional variables that might help
explain variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and local solar hour (for diurnal effects). Solar hour was calculated using the
{solaR} package [@perpinan2012jss] within the R statistical environment [@rcoreteam2019]
using the daily locations. We also compiled several meteorological variables shown to
affect haul-out behavior in other Arctic pinnipeds [@reder2003pb; @udevitz2009cjz;
@perry2017]. In particular, we linked locations from satellite tags to key weather values
from the North American Regional Reanalysis (NARR) model produced by the National Centers
for Environmental Prediction [@mesinger2006bams]. The NARR model assimilates observational
data to produce a long-term picture of weather over North America. Numerous weather
variables are made available across the region 8 times daily. For this study, NARR weather
values were subset to the extent of our study area over the Bering and Chukchi Seas at
3-hr intervals based on the native grid resolution of 32 km (1024 sq. km). The following
meteorological variables were interpolated and assigned to seal locations using a bilinear
method: 1) air temperature at 2m above the earth's surface, 2) wind consisting of
northerly and easterly vector components converted to wind speed using the Euclidean norm,
3) barometric pressure at sea level, and 4) precipitation (Table
\@ref(tab:tblCovariates)).

@@

To properly account for temporal dependence within a computationally tractable framework,
we used generalized linear mixed pseudo-models (GLMPMs; [@verhoef2010cs]) to model
variation in haul-out behavior as a function of (1) covariate predictors, (2) temporally
autocorrelated random effects, and (3) individual random effects representing
heterogeneity in individual behavior. We used the {glmmLDTS} package [@verhoef2010cs] to
implement GLMPMs in the R programming environment [@rcoreteam2021]. We explored two
different model formulations for our data, and owing to the large number of records, we
fit separate models to bearded, ribbon, and spotted seal data sets.  In our first model
formulation, for each species, we fitted a year-independent model that predicted average
haul-out behavior as a function of demographic, environmental, seasonal, and diurnal
effects. Second, for ribbon and spotted seals (which had considerably more data than
bearded seals), we fitted models that included all the effects from the first model, but
also permitted annual variation in haul-out timing. This second class of models was used
to examine whether haul-out patterns varied by year and to determine the annual timing of
apparent peaks in haul-out behavior.

@@

The second set of models that included annual variation in haul-out patterns uncovered
significant contributions for linear and quadratic interactions between day and year in
both ribbon seals (day $\times$ year,
$F_{`r paste(ribbon_year_type3$day_year$Num.df, ribbon_year_type3$day_year$Den.df, sep=",")`}$
= `r ribbon_year_type3$day_year$F.value`; $p$ = `r ribbon_year_type3$day_year$Prob.F`;
day^2^:year,
$F_{`r paste(ribbon_year_type3$day2_year$Num.df, ribbon_year_type3$day2_year$Den.df, sep=",")`}$
= `r ribbon_year_type3$day2_year$F.value`; $p$ = `r ribbon_year_type3$day2_year$Prob.F`)
and spotted seals (day $\times$ year,
$F_{`r paste(spotted_year_type3$day_year$Num.df, spotted_year_type3$day_year$Den.df, sep=",")`}$
= `r spotted_year_type3$day_year$F.value`; $p$ = `r spotted_year_type3$day_year$Prob.F`;
day^2^:year,
$F_{`r paste(spotted_year_type3$day2_year$Num.df, spotted_year_type3$day2_year$Den.df, sep=",")`}$
= `r spotted_year_type3$day2_year$F.value`; $p$ =
`r spotted_year_type3$day2_year$Prob.F`). Predicted distributions of haul-out activity
were largely unimodal, but varied some among and within years with respect to the timing
and magnitude of haul-out peaks (Figure \@ref(fig:annualHO)). It is important to note that
predicted variation in annual haul-out patterns likely reflects both process error and
sampling variability. While we did remove any years where only one deployment in a species
+ age-sex group was present, there are still some years where the pattern shown is
informed by a small number of individuals that may not represent population level processes.
